----
title: "Preliminary Exploratory Data Analysis" 
subtitle: "Class Project"                     
author: "Anonymous"                        
date: "11/17/2024"                             
output:                                        
  html_notebook:                                 
    fig_height: 7                               
    fig_width: 10                                  
----
# Data Sources

- Golf Dataset (golf_data):
  - Player_initial_last: Player names.
  - strokes: Total strokes (a key performance metric).
  - hole_FDP: Hole performance (points).
  - sg_putt, sg_t2g, sg_total: Additional performance metrics for putting, tee-to-green, and total score.

- Basketball Dataset (basketball_data):
  - player: Player names.
  - age: Age of the player.
  -experience: Years of professional experience.

----

# Setup

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)

# Define file paths for datasets
golf_path <- "/Users/daniel/Desktop/stat 184/ASA All PGA Raw Data - Tourn Level.csv"
basketball_path <- "/Users/daniel/Desktop/stat 184/Player Season Info.csv"

# Load the data into R
golf_data <- read.csv(golf_path)
basketball_data <- read.csv(basketball_path)
```

```{r}
# Clean player names in both datasets
golf_data <- golf_data %>%
  mutate(Player_initial_last = str_trim(Player_initial_last))  # Remove whitespace

basketball_data <- basketball_data %>%
  mutate(player = str_trim(player))  # Remove whitespace
```

#Data Wrangling
#1. General data wrangling using various data verbs like filter, mutate, summarise, arrange, group_by, etc.
```{r}

golf_clean <- golf_data %>%
  filter(!is.na(hole_FDP), !is.na(strokes), strokes > 0) %>% 
  mutate(
    ScoringEfficiency = hole_FDP / strokes,  # Create a new efficiency metric
    PerformanceCategory = if_else(ScoringEfficiency >= 1, "High", "Low")
  ) %>%
  group_by(PerformanceCategory) %>%
  summarise(
    avg_hole_FDP = mean(hole_FDP, na.rm = TRUE),
    avg_strokes = mean(strokes, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_hole_FDP))  # Arrange by performance
golf_clean


```

#2 Joins: Merging Golf and Basketball Data
```{r}
# Join datasets on player names
combined_data <- inner_join(golf_data, basketball_data, by = c("Player_initial_last" = "player"))
head(combined_data)



```

#3Wide to Long Transformation
```{r}
# Transform golf data from wide to long format
golf_long <- golf_data %>%
  select(Player_initial_last, sg_putt, sg_t2g, sg_total) %>%
  pivot_longer(cols = c(sg_putt, sg_t2g, sg_total), names_to = "Metric", values_to = "Value")

head(golf_long)

```

#4 Regular Expressions
```{r}
# Extract initials from basketball player names
basketball_clean <- basketball_data %>%
  mutate(Initials = str_extract(player, "\\b[A-Z]"))  # Extract the first capital letter
head(basketball_clean)

```

#5 reduction and/or transformation functions like mean, sum, max, min, n(), rank, pmin, etc.
```{r}
# Summarize basketball data
basketball_summary <- basketball_data %>%
  summarise(
    avg_age = mean(age, na.rm = TRUE),
    max_experience = max(experience, na.rm = TRUE),
    total_players = n()
  )
basketball_summary

```

#Visualizations

##Golf: Scoring Efficiency Distribution
1. Relationship Between strokes, hole_FDP, and ScoringEfficiency

```{r}
ggplot(golf_clean, aes(x = avg_strokes, y = avg_hole_FDP, color = PerformanceCategory)) +
  geom_point(size = 4, alpha = 0.7) +
  labs(title = "Strokes vs. Hole FDP by Performance Category",
       x = "Average Strokes",
       y = "Average Hole FDP",
       color = "Performance Category") +
  theme_minimal()



```
2. Distribution of ScoringEfficiency by PerformanceCategory
```{r}

ggplot(golf_clean, aes(x = PerformanceCategory, y = avg_hole_FDP, fill = PerformanceCategory)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  labs(title = "Average Hole FDP by Performance Category",
       x = "Performance Category",
       y = "Average Hole FDP",
       fill = "Performance Category") +
  theme_minimal()

```
##Basketball Data
1.Relationship Between age, experience, and PointsPerExperience
```{r}
basketball_clean <- basketball_data %>%
  mutate(PointsPerExperience = age / experience)

ggplot(basketball_clean, aes(x = experience, y = age, color = PointsPerExperience)) +
  geom_point(alpha = 0.7) +
  labs(title = "Age vs. Experience vs. Points Per Experience",
       x = "Experience (Years)",
       y = "Age",
       color = "Points Per Experience") +
  theme_minimal()

```

2. Points Per Experience by Age Group
```{r}
basketball_clean <- basketball_clean %>%
  mutate(AgeGroup = case_when(
    age < 25 ~ "Young",
    age >= 25 & age < 30 ~ "Prime",
    age >= 30 ~ "Veteran"
  ))

ggplot(basketball_clean, aes(x = AgeGroup, y = PointsPerExperience, fill = AgeGroup)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Points Per Experience by Age Group",
       x = "Age Group",
       y = "Points Per Experience",
       fill = "Age Group") +
  theme_minimal()

```




#Insights and Summary
- Techniques Used:
  General Data Wrangling: Applied functions such as filter, mutate, group_by, and summarise to clean and            organize the data.
  Joins: Combined the golf and basketball datasets based on player names.
  Pivoting: Reshaped the golf data from wide format to long format using pivot_longer.
  Regular Expressions: Extracted players' initials from their names in the basketball dataset.
  Reduction/Transformation Functions: Calculated summary metrics using functions like mean, max, and n().
  
- Key Insights:
Golf:
Players with lower stroke counts tend to achieve higher hole_FDP scores and better ScoringEfficiency metrics.
Categorizing player performance into "High" and "Low" groups offers meaningful distinctions between top and underperforming players.

Basketball:
Players with greater experience consistently show higher performance when evaluated by points per year of experience.
Younger players, while often showing potential, exhibit greater variability in their performance compared to veterans.
